---
title: "Weekend effect"
output: html_document
date: "2024-11-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Weekend effect on ICU mortality

<i><b>To investigate whether ICU admissions over the weekend increase the risk of mortality. </i></b><br> - Data Extraction: Admissions will be categorized into weekday (Monday-Friday) and weekend (Saturday-Sunday) groups.<br> - Inclusion/Exclusion Criteria: All ICU admissions, regardless of patient age, will be included. Patients with incomplete data on admission time or those with transfers between ICUs will be excluded.<br> - EDA: Explore trends in mortality across weekday and weekend admissions and adjust for confounding variables like illness severity, age, and comorbidities. I can use propensity matching score to adjust for these confounding factors and match the similar cases to each other.<br> - Software: R for data analysis and survival analysis (using the survival library).

```{r library, include=FALSE}
# Load dataset
library(readr)
library(dplyr)
```

```{r dataset}
#Define the columns to be converted to factors
cat_cols <- c("admission_type", "admission_location", "discharge_location", "insurance", "language", "religion", "marital_status", "ethnicity", "diagnosis")

# Define the columns to be converted to logic
logic_cols <- c("hospital_expire_flag", "has_chartevents_data")

adm <- readr::read_csv("../mimic_data/admissions.csv") %>%
  mutate(across(all_of(cat_cols), as.factor)) %>%
  mutate(across(all_of(logic_cols), as.logical))
# 58976 obs of 19 vars

pt_icu <- readr::read_csv("../mimic_data/pt_icu_outcome.csv")
patients <- readr::read_csv("../mimic_data/patients.csv")
vasopressors <- readr::read_csv("../mimic_data/vasopressors.csv")
output <- readr::read_csv("../mimic_data/output_hourly.csv")

length(unique(vasopressors$icustay_id)) #8379
length(unique(output$icustay_id)) #53876

```

-   Add admission and discharge day in the week (Monday to Sunday)
-   `hospital_expire_flag`: logical variable
-   `admission_type`, `admission_location`, `discharge_location`, `insurance`, `language`, `religion`, `marital_status`, `ethnicity`: factor variable

## Pre-processing data notes:

-   Admission type: remove NEWBORN and ELECTIVE because newborn is out of scope and elective is planned beforehand\<br.
-   Remove TRANSFER from other hospital or nursing home because these patients are more likely to be more ill than whom being admitted straight in the emergency<br>
-   I can also calculate edouttime - edregtime to see how long the patients stay in ED
-   I will stratify age according to the ABS age bracket: <https://www.abs.gov.au/statistics/standards/age-standard/2014-version-17>

```{r preprocess, echo=FALSE}
# Pre-processing adm dataset
adm <- adm %>%
  filter(!(admission_type %in% c("NEWBORN", "ELECTIVE"))) %>% # 43407 obs
  filter(admission_location == "EMERGENCY ROOM ADMIT") %>%
  mutate(admission_type = droplevels(admission_type),
         admission_location = droplevels(admission_location)
         )
# 34660 obs

adm <- adm %>%
  mutate(
    # Convert admittime and dischtime to date-time format
    admittime = as.POSIXct(admittime, format = "%Y-%m-%d %H:%M:%OS"),
    dischtime = as.POSIXct(dischtime, format = "%Y-%m-%d %H:%M:%OS"),
    deathtime = as.POSIXct(deathtime, format = "%Y-%m-%d %H:%M:%S"),
    # Extract and convert admission and discharge days to factors
    admit_day = as.factor(weekdays(admittime)),
    disch_day = as.factor(weekdays(dischtime))
  ) %>%
  distinct()

# Prepare for cox proportional hazard ratio
# Create 'time_to_event' column
adm <- adm %>%
  mutate(
    time_to_event = as.numeric(
      pmin(dischtime, deathtime, na.rm = TRUE) - admittime, # Time difference in seconds
      units = "days" # Convert to days
    ),
    event = ifelse(!is.na(deathtime), 1, 0) # 1 = died, 0 = censored
  )

# Select columns in patients and pt_icu
pt_icu_selected <- pt_icu %>%
  select(subject_id, age_years) %>%
  distinct(subject_id, .keep_all = TRUE)

patients_selected <- patients %>%
  select(subject_id, gender) %>%
  distinct(subject_id, .keep_all = TRUE)

library(ggplot2)
library(dplyr)
library(lubridate)

library(tidyverse)
library(tidytext)

adm <- adm %>%
  mutate(diagnosis = as.character(diagnosis))

# Process the diagnosis column to extract word frequencies
word_freq <- adm %>%
  # Tokenize words, separating them by space, punctuation, and delimiters like ";"
  unnest_tokens(word, diagnosis) %>%
  # Remove common stopwords
  anti_join(stop_words) %>%
  # Filter out NA, numbers, and meaningless punctuation
  filter(!str_detect(word, "\\d"), !str_detect(word, "[[:punct:]]")) %>%
  # Count word occurrences
  count(word, sort = TRUE) %>%
  # Calculate percentage of each word
  mutate(percentage = n / sum(n) * 100)

# View the word frequencies
print(word_freq)

# Create a bubble chart for top 10 most frequent words
ggplot(word_freq %>% slice_max(n, n = 10), aes(x = reorder(word, n), y = percentage, size = n, fill = n)) +
  geom_point(alpha = 0.7, shape = 21) +
  labs(
    title = "Most Frequently Appearing Words in Diagnoses",
    x = "Words",
    y = "Percentage (%)"
  ) +
  scale_size(range = c(3, 15), name = "Frequency") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold")
  )

library(wordcloud2)

# Generate a word cloud
wordcloud2(
  data = word_freq %>% slice_max(n, n = 50), 
  size = 0.5,                  # Increase or decrease size (default is 1)
  color = 'random-dark', 
  shape = 'cloud'
)

# # Aggregate the count of admissions by day of the week
# adm_summary <- adm %>%
#   group_by(admit_day) %>%
#   summarise(count = n())
#   
#   adm_summary
# # Create the bar plot for admissions by day of the week
# ggplot(adm_summary, aes(x = admit_day, y = count, fill = admit_day)) +
#   geom_bar(stat = "identity", color = "black") +
#   labs(
#     title = "Admissions by Day of the Week",
#     x = "Day of the Week",
#     y = "Number of Admissions"
#   ) +
#   theme_minimal() +
#   theme(
#     axis.text.x = element_text(angle = 45, hjust = 1),
#     plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
#     legend.position = "none"
#   )


```

```{r merge, echo = FALSE}
# Merge all dataset
merge_df <- adm %>% # 34733 records
  left_join(patients_selected, by = "subject_id") %>%
  left_join(pt_icu_selected, by = "subject_id") %>%
  distinct(hadm_id, .keep_all = TRUE) %>% # all the hadm_id are currently unique
  mutate(gender = as.factor(gender)) %>%
  filter(age_years != 0) %>% # 34553 records
  mutate(
    age_group = as.factor(case_when(
      age_years <= 24 ~ "Young",
      age_years > 24 & age_years <= 44 ~ "Adult",
      age_years > 44 & age_years <= 64 ~ "Middle age",
      age_years > 64 ~ "Senior",
      TRUE ~ "Unknown"  # For any missing or undefined age values
    ))
  ) %>%
  mutate(ethnicity = case_when(
    ethnicity %in% c("WHITE", "WHITE - RUSSIAN", "WHITE - OTHER EUROPEAN", "WHITE - BRAZILIAN", "WHITE - EASTERN EUROPEAN") ~ "White",
    ethnicity %in% c("BLACK/AFRICAN AMERICAN", "BLACK/AFRICAN", "BLACK/HAITIAN", "BLACK/CAPE VERDEAN") ~ "Black",
    ethnicity %in% c("ASIAN", "ASIAN - VIETNAMESE", "ASIAN - CHINESE", "ASIAN - FILIPINO", "ASIAN - CAMBODIAN", "ASIAN - KOREAN", "ASIAN - JAPANESE", "ASIAN - THAI", "ASIAN - ASIAN INDIAN", "ASIAN - OTHER") ~ "Asian",
    ethnicity %in% c("HISPANIC OR LATINO", "HISPANIC/LATINO - GUATEMALAN", "HISPANIC/LATINO - PUERTO RICAN", "HISPANIC/LATINO - SALVADORAN", "HISPANIC/LATINO - DOMINICAN", "HISPANIC/LATINO - CENTRAL AMERICAN (OTHER)", "HISPANIC/LATINO - COLOMBIAN", "HISPANIC/LATINO - HONDURAN", "HISPANIC/LATINO - CUBAN", "HISPANIC/LATINO - MEXICAN") ~ "Hispanic",
    TRUE ~ "Other"  # Catch-all for any other categories
  )) %>%
  mutate(ethnicity = as.factor(ethnicity)) %>%  # Convert to factor
  mutate(
    weekend_admit = as.factor(case_when(
      admit_day %in% c("Saturday", "Sunday") ~ "Yes",
      TRUE ~ "No")
    )) %>%
  mutate(marital_status = case_when(
    marital_status %in% c("LIFE PARTNER", "MARRIED") ~ "MARRIED",
    marital_status %in% c("UNKNOWN (DEFAULT)") ~ "UNKNOWN",
    is.na(marital_status) ~ "UNKNOWN",
    TRUE ~ marital_status
  )
  )
#34480 records
summary(merge_df)

# Check age_years: Filter records where age_years is 0 -> 160 records
age_zero_records <- merge_df %>%
  filter(age_years == 0)

#unique(merge_df$ethnicity)

```

```{r test, echo = FALSE}
library(dplyr)
library(ggplot2)

# Filter for deaths
death_data <- merge_df %>% 
  filter(hospital_expire_flag == 1)

# Calculate percentages for each diagnosis
top_diagnoses_percentage <- death_data %>%
  group_by(weekend_admit, diagnosis) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(weekend_admit) %>%
  mutate(percentage = (count / sum(count)) * 100) %>% # Convert counts to percentages
  arrange(desc(percentage)) %>%
  slice_max(percentage, n = 10) %>% # Get top 10 diagnoses by percentage for each group
  ungroup()

# Pivot data for plotting
top_diagnoses_long <- top_diagnoses_percentage %>%
  mutate(adm_day_type = ifelse(weekend_admit == "Yes", "Weekend", "Weekday")) %>%
  select(adm_day_type, diagnosis, percentage)

# Plot comparison of diagnoses as percentages
p <- ggplot(top_diagnoses_long, aes(x = reorder(diagnosis, percentage), y = percentage, fill = adm_day_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(
    title = "Top diagnoses contributing to deaths weekend and weekday admission",
    x = "Diagnosis",
    y = "Percentage (%)",
    fill = "Admission Type"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
  )


```

```{r model}
# Remove unnecessary columns in merge_df -> clean_df
# Select relevant columns in merge_df -> select_df
select_df <- merge_df %>%
  select(age_group, age_years, gender, hospital_expire_flag, weekend_admit, admit_day, ethnicity, insurance, marital_status, time_to_event, event)
  
  
library(compareGroups)
# Run the compareGroups analysis
comp <- compareGroups(weekend_admit ~ ., data = select_df)

# Create the comparison table
compareGroups::createTable(comp)

```

```{r model1}
library(MatchIt)
library(forcats)
library(cobalt)

sum(is.na(select_df$insurance))
# Propensity score matching using coarsened exact matching (CEM)
match1 <- matchit(weekend_admit ~ age_group,
                  method = 'cem',
                  data = select_df)
# Based on gender
match2 <- matchit(weekend_admit ~ gender,
                  method = 'cem',
                  data = select_df)
# Based on ethnicity
match3 <- matchit(weekend_admit ~ ethnicity,
                  method = 'cem',
                  data = select_df)
# Based on marital_status
match4 <- matchit(weekend_admit ~ marital_status,
                  method = 'cem',
                  data = select_df)
# Based on insurance
match5 <- matchit(weekend_admit ~ insurance,
                  method = 'cem',
                  data = select_df)
# Based on everything
match6 <- matchit(weekend_admit ~ age_group + gender + ethnicity,
                  method = 'cem',
                  data = select_df)
summary(match1)
summary(match2)
summary(match3)
summary(match4)
summary(match5)
summary(match6)

md1 <- match.data(match1)
md2 <- match.data(match2)
md3 <- match.data(match3)
md4 <- match.data(match4)
md5 <- match.data(match5)
md6 <- match.data(match6)

love.plot(match1)
love.plot(match2)
love.plot(match3)
love.plot(match4)
love.plot(match5)
love.plot(match6)

# Use an appropriate linear model to estimate the effect of weekend admission on mortality in (i) the raw data and (ii) the matched data.
# Logistic regression model on raw data
raw_model <- glm(hospital_expire_flag ~ weekend_admit + age_group, data = select_df, family = binomial)
summary(raw_model)

raw_model6 <- glm(hospital_expire_flag ~ weekend_admit + age_group + gender + ethnicity, data = select_df, family = binomial)
summary(raw_model6)

# Logistic regression model on matched data
matched_model1 <- glm(hospital_expire_flag ~ weekend_admit + age_group, data = md1, family = binomial)
summary(matched_model1)

matched_model2 <- glm(hospital_expire_flag ~ weekend_admit + gender, data = md2, family = binomial)
summary(matched_model2)

matched_model3 <- glm(hospital_expire_flag ~ weekend_admit + age_group + gender, data = md3, family = binomial)
summary(matched_model3) # Best model so far AIC 12571

matched_model4 <- glm(hospital_expire_flag ~ weekend_admit + age_group + gender + ethnicity, data = md4, family = binomial)
summary(matched_model4)

matched_model5 <- glm(hospital_expire_flag ~ weekend_admit + insurance, data = md5, family = binomial)
summary(matched_model5)

matched_model6 <- glm(hospital_expire_flag ~ weekend_admit + age_group + gender + ethnicity, data = md6, family = binomial)
summary(matched_model6)
```

## Comments:

-   There is a weekend effect based on age and gender, not so much with ethnicity (although there are some significant difference in Black, Hispanic and Other ethnicity but not consistent)
-   Interpret the number above about the model after matching (adjusted for age and gender)
-   Play around with different matching method: to looking into different confounding factors that may interfere with the validity of the weekend effect result. This can be done later in 2 weeks time
-   These data are enough to document into my draft now
-   Some ideas: look into the marital status later to see if there is anything interesting. Extract the diagnosis to see which one is the most prevalence in terms of words, can do something like bubble chart for words. Time series analysis to look at the trend throughout many weeks for different groups

```{r cox}


# Load required libraries
library(survival)
library(dplyr)
library(survminer)

# Prepare dataset for Cox model
cox_data <- select_df %>%
  filter(!is.na(time_to_event) & !is.na(hospital_expire_flag)) %>%
  mutate(event = hospital_expire_flag)

# Fit Cox proportional hazards model
cox_model <- coxph(Surv(time_to_event, event) ~ weekend_admit, data = as.data.frame(cox_data))
cox_model1 <- coxph(Surv(time_to_event, event) ~ weekend_admit + age_group + gender + ethnicity, data = as.data.frame(cox_data))
cox_model2 <- coxph(Surv(time_to_event, event) ~ weekend_admit + age_group + gender + ethnicity + insurance + marital_status, data = as.data.frame(cox_data))

# View summary of the model
summary(cox_model)
summary(cox_model1)
summary(cox_model2)

# Hazard ratios with confidence intervals
exp(coef(cox_model))  # Hazard ratios
exp(confint(cox_model))  # Confidence intervals

# Visualize hazard ratios with ggforest
ggforest(cox_model1)

# Fit a Cox model with stratification for age_group and ethnicity
cox_model_strata <- coxph(Surv(time_to_event, event) ~ weekend_admit + gender + strata(age_group, ethnicity), data = cox_data)

# Summary of the model
summary(cox_model_strata)

# Check Proportional Hazards Assumption
ph_test <- cox.zph(cox_model1)
print(ph_test)
plot(ph_test)
```

```{r km}
library(survival)
library(survminer)

# Create a survival object
surv_obj <- Surv(time = select_df$time_to_event, event = select_df$hospital_expire_flag)

# Fit Kaplan-Meier survival curves
km_fit <- survfit(surv_obj ~ weekend_admit, data = select_df)

# Plot survival curves
ggsurvplot(
  km_fit,
  data = select_df,
  conf.int = TRUE,                  # Show confidence intervals
  pval = TRUE,                      # Display p-value for log-rank test
  legend.labs = c("Weekday", "Weekend"),  # Labels for legend
  legend.title = "Admission Type",
  xlab = "Time to Event (Days)",    # X-axis label
  ylab = "Survival Probability",    # Y-axis label
  palette = c("#E69F00", "#56B4E9"), # Custom colors for curves
  risk.table = TRUE,                # Add risk table below the plot
  ggtheme = theme_minimal()         # Apply a minimal theme
)

```
